---
title: 'DATA 698: Masters Research Project'
author: "Gabriella Martinez"
date: "Last edited `r format(Sys.time(), '%B %d, %Y')`"
output:
      html_document:
        toc: yes
        toc_float: yes
        theme: yeti
        highlight: kate
        font-family: "Arial"
        code_folding: show
      html_notebook: default
      pdf_document:
        latex_engine: xelatex
      geometry: left=0.5cm,right=0.5cm,top=1cm,bottom=2cm
urlcolor: blue
---


# Packages
```{r warning=FALSE, message=FALSE, results='hide'}
#load libraries
library(tidyverse)
library(tidycensus)
library(rvest)

```
```{r echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
census_api_key("8ae64973afa0ae475a345905a9db7a4b3f06e783", overwrite = TRUE, install = TRUE)
```


# Load Data
## Election Data
```{r warning=FALSE}
#https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/VOQCHQ
elections <- read_csv("https://raw.githubusercontent.com/gabbypaola/DATA698/refs/heads/main/data/countypres_2000-2020.csv")

#glimpse(elections)
```

### Cleaning Election data
```{r}
#identify empty and NA values. 57 NA values in the county_fips column
colSums(elections == "" | is.na(elections))

elections %>% 
  filter(is.na(county_fips))

elections %>% 
  filter(state_po=="DC")
```

```{r}
#clean elections data
elections_data <- elections %>% 
  #new name = old name
  rename(state_abbr = state_po, pol_identity = party, FIPS = county_fips) %>% 
  mutate(FIPS = ifelse(state_abbr == "DC", "11001", FIPS))

#there are 52 NAs remaining
remaining_nas <- elections_data %>% 
  filter(is.na(FIPS)) 

#what are the NAs? 
#ME MAINE UOCAVA= Uniformed Service & Overseas (UOCAVA) Voters- https://www.maine.gov/sos/cec/elec/voter-info/uocava.html
#CT= STATEWIDE WRITEIN ~ You may decide to write in the name of a candidate who is not listed on your ballot- https://www.usa.gov/write-in-candidates
#RI Federal Precinct- no info found
remaining_nas %>% 
  count(state_abbr, county_name) 
```

The remaining NA values in the FIPS column are votes assigned at a state-wide level, not to any count. The "MAINE UOCAVA" county record for the state of Maine represents the count of votes from Uniformed Service & Overseas (UOCAVA) Voters. The "STATEWIDE WRITEIN" for Connecticut represents the count of votes for self-selected candidates not on the presidential ballot. It is unclear what the "FEDERAL PRECINCT" for the state of Rhode Island exactly represents. Either way, our analysis will be conducted at the county level, so these records cannot be used. 

Next we will assess the effect that removing these votes will have on our overall analysis.

```{r}
#nas
nrow(remaining_nas)

# Determine the total number of records in the table.
nrow(elections_data)

round(nrow(remaining_nas)/nrow(elections_data)*100,3)
```

```{r}
# Determine the total number of votes cast across all counties in all elections.
votecount <- elections_data %>% 
  summarise(count= sum(candidatevotes))

votecount
```

```{r}
# Determine how many votes are associated with state-level counts
null_fips_votecount <- remaining_nas %>% 
  summarise(count=sum(candidatevotes))

null_fips_votecount

round((null_fips_votecount$count/votecount$count)*100,3)
```

There were 52 records with state-level counts and null FIPS values in the data, representing 13009 votes.
This amounts to 0.072% of the total records and 0.002% of the total votes.

The records with state-level counts and null FIPS values represent a small percentage of the total, and they are unlikely to change the overall analysis. Given our assessment, the records will be removed.

```{r}
#transform data- drop NAs, keep dem and gop only, group records for each candidate by county and year
candidate_votes <- elections_data %>% 
  filter(!is.na(FIPS), pol_identity %in% c('DEMOCRAT', 'REPUBLICAN')) %>% 
  group_by(FIPS,county_name, state, candidate, year, pol_identity, totalvotes) %>% 
  summarise(candidate_votes = sum(candidatevotes)) %>% 
  ungroup()

#spread the candidate votes values
elections_pivot_df <- candidate_votes %>% 
   pivot_wider(id_cols = c(year, FIPS, county_name, state, totalvotes),
               names_from = pol_identity,
               values_from = candidate_votes) %>% 
  rename(votes_dem = DEMOCRAT,  votes_gop = REPUBLICAN
         #votes_other = OTHER,votes_grn = GREEN, votes_lib = LIBERTARIAN
         )
```


## Census Bureau data

About Census Bureau American Community Survey (ACS) data
https://www.census.gov/programs-surveys/acs/guidance/estimates.html

### Citizen Voting Age Population
Citizen Voting Age Population, Census Bureau population estimates generated using the American Community Survey
```{r, message=FALSE, warning=FALSE, results='hide'}
#CVAP- Citizen Voting Age Population, Census Bureau population estimates generated using the American Community Survey

#https://www.census.gov/programs-surveys/decennial-census/about/voting-rights/cvap.2010.html#list-tab-1518558936 (2008)
cvap2008 <- read_csv("https://raw.githubusercontent.com/GitableGabe/DATA_698/refs/heads/main/data/CountyCVAP_2006-2010.csv?token=GHSAT0AAAAAACXYKDAYQCHUVJY2V6BVWU7SZXPAZJQ") %>% 
  rename_with(tolower) %>% 
  mutate(year=2008)

#https://www.census.gov/programs-surveys/decennial-census/about/voting-rights/cvap.2014.html#list-tab-1518558936 (2012)
cvap2012 <- read_csv("https://raw.githubusercontent.com/GitableGabe/DATA_698/refs/heads/main/data/CountyCVAP_2010-2014.csv?token=GHSAT0AAAAAACXYKDAYHOL27SGWSEL2AS6IZXPAYSQ") %>% 
  rename_with(tolower) %>% 
  mutate(year=2012)

#https://www.census.gov/programs-surveys/decennial-census/about/voting-rights/cvap/2014-2018-CVAP.html (2016)
cvap2016 <- read_csv("https://raw.githubusercontent.com/GitableGabe/DATA_698/refs/heads/main/data/CountyCVAP_2014-2018.csv?token=GHSAT0AAAAAACXYKDAZJU7ABMJMRNP5WOSIZXPATUQ") %>% 
  mutate(year=2016)

#https://www.census.gov/programs-surveys/decennial-census/about/voting-rights/cvap/2017-2021-CVAP.html (2020)
cvap2020 <- read_csv("https://raw.githubusercontent.com/GitableGabe/DATA_698/refs/heads/main/data/CountyCVAP_2017-2021.csv?token=GHSAT0AAAAAACXYKDAYJWVR6SZPSH4NRMSSZXPASSQ") %>% 
  mutate(year=2020) 

cvap_df <- rbind(cvap2008, cvap2012, cvap2016, cvap2020) %>% 
  filter(lntitle == 'Total', !str_detect(geoname, "Puerto Rico")) %>% 
  mutate(FIPS = str_sub(geoid, -5)) %>% 
  select(c('year', 'FIPS', 'geoname', 'cvap_est'))

#identify empty and NA values
colSums(cvap_df == "" | is.na(cvap_df))
```

#### Merge with Election data

```{r}
voting_info_df <- left_join(elections_pivot_df, cvap_df, by = c("FIPS", "year"))

voting_info_df
```

```{r}
#identify empty and NA values
colSums(voting_info_df == "" | is.na(voting_info_df))

voting_info_df_NAs <- voting_info_df %>% 
  filter(is.na(geoname), is.na(cvap_est))

voting_info_df_NAs

unique(voting_info_df_NAs$year)
```


```{r}
voting_info_df <- voting_info_df %>% 
  filter(year >= 2008)

voting_info_df_NAs2 <- voting_info_df %>% 
  filter(is.na(geoname), is.na(cvap_est))

voting_info_df_NAs2
```

```{r}
voting_info_df <- voting_info_df %>% 
  filter(state != "ALASKA")

voting_info_df_NAs3 <- voting_info_df %>% 
  filter(is.na(geoname), is.na(cvap_est))

voting_info_df_NAs3
```

```{r}
voting_info_df_cleaning <- voting_info_df %>% 
  filter(FIPS %in% c('29095', '36000', '51019', '51515')) %>% 
  arrange(year, FIPS)

voting_info_df_cleaning
```

```{r}
voting_info_df_cleaning %>% 
  count(FIPS, state, county_name, geoname) %>% 
  filter(geoname %in% c("Jackson County, Missouri", "Bedford County, Virginia")) %>% 
  select(-n)
```


```{r}
# Define the counties to filter and group data by year and state
county_groups <- voting_info_df %>%
  filter(FIPS %in% c('29095', '36000', '51019', '51515')) %>%
  group_by(year, state) %>%
  summarise(    # Concatenate FIPS codes and county names
    FIPS = paste(unique(FIPS), collapse = ", "),
    county_name = paste(unique(county_name), collapse = ", "),
            across(where(is.numeric), sum, na.rm = TRUE)) %>% 
  mutate(geoname = case_when(state == "MISSOURI" ~ "Jackson County, Missouri",
                             state == "VIRGINIA" ~ "Bedford County, Virginia"))

county_groups
```


```{r}
#remove the previous observations
voting_info_df <- voting_info_df %>% 
  filter(!FIPS %in% c('29095', '36000', '51019', '51515'))

#replace with the calculated observations
voting_info_df <- rbind(voting_info_df, county_groups)

```


#### Calculate additional columns

```{r}
voting_info_final <- voting_info_df %>% 
  mutate(#voters who did not choose the Democratic or Republican party
         votes_other = totalvotes - votes_dem - votes_gop,
         #voter share attributes
         voter_share_major_party = (votes_dem + votes_gop) / totalvotes,
         voter_share_dem = votes_dem/totalvotes,
         voter_share_gop = votes_gop/totalvotes,
         voter_share_other = votes_other/totalvotes,
         #raw differences
         rawdiff_dem_vs_gop = votes_dem - votes_gop,
         rawdiff_gop_vs_dem = votes_gop - votes_dem,
         rawdiff_dem_vs_other = votes_dem - votes_other,
         rawdiff_gop_vs_other = votes_gop - votes_other,
         rawdiff_other_vs_dem = votes_other - votes_dem,
         rawdiff_other_vs_gop = votes_other - votes_gop,
         #percentage difference
         pctdiff_dem_vs_gop = (votes_dem - votes_gop) / totalvotes,
         pctdiff_gop_vs_dem = (votes_gop - votes_dem) / totalvotes,
         pctdiff_dem_vs_other = (votes_dem - votes_other) / totalvotes,
         pctdiff_gop_vs_other = (votes_gop - votes_other) / totalvotes,
         pctdiff_other_vs_dem = (votes_other - votes_dem) / totalvotes,
         pctdiff_other_vs_gop = (votes_other - votes_gop) / totalvotes,
         # get winning political party
         winning_party = case_when(votes_dem > votes_gop & votes_dem > votes_other ~ "Democratic Party",
                                   votes_gop > votes_dem & votes_gop > votes_other ~ "Republican Party",
                                   votes_other > votes_dem & votes_other > votes_gop ~ "Other Party"),
         pct_margin_of_victory = case_when(winning_party == "Democratic Party" ~ round(((votes_dem - votes_gop) / totalvotes)*100,3), #votes_dem > votes_gop
                                          winning_party == "Republican Party" ~ round(((votes_gop - votes_dem) / totalvotes)*100,3), #votes_gop > votes_dem
                                          )
         ) 

glimpse(voting_info_final)
```

### Educational Attainment data

```{r, warning= FALSE, message=FALSE, results='hide'}
# To obtain data for the 2008 population from the American Community Survey (ACS), 
# you should use the 2006-2008 ACS 3-Year Estimates. This dataset aggregates data 
# collected over those three years, providing insights for the population during that period.
# 5 year ACS data unavailable for 2008. 3 year ACS data was discontinued after 2009.

#load 2008 data using API
ed_attain2008 <- get_acs(
  geography = "county",
  variables = c(paste0("B15001_00", seq(01,09),"E"),paste0("B15001_0", seq(10,83),"E")),
  year = 2008,
  survey = "acs3",
  cache_table = TRUE) %>% 
  mutate(year=2008)

#2012 data and onward uses the 5 year ACS data
#load 2012 data using API
ed_attain2012 <- get_acs(
  geography = "county",
  variables = c(paste0("B15001_00", seq(01,09),"E"),paste0("B15001_0", seq(10,83),"E")),
  year = 2012,
  survey = "acs5",
  cache_table = TRUE) %>% 
  mutate(year=2012)

#load 2016 data using API
ed_attain2016 <- get_acs(
  geography = "county",
  variables = c(paste0("B15001_00", seq(01,09),"E"),paste0("B15001_0", seq(10,83),"E")),
  year = 2016,
  survey = "acs5",
  cache_table = TRUE) %>% 
  mutate(year=2016)

#load 2020 data using API
ed_attain2020 <- get_acs(
  geography = "county",
  variables = c(paste0("B15001_00", seq(01,09),"E"),paste0("B15001_0", seq(10,83),"E")),
  year = 2020,
  survey = "acs5",
  cache_table = TRUE) %>% 
  mutate(year=2020)

```
#### Get column names

```{r}
#check column names
#get column names 2008
url08 <- "https://api.census.gov/data/2008/acs/acs3/groups/B15001.html"

webpage08 <- read_html(url08)

table08 <- webpage08 %>%
  html_node("table") %>%  # Adjust the selector if necessary
  html_table() %>% 
  select(c("Name","Label","Concept","Required","Attributes","Limit","Predicate Type","Group"))

filteredtable08 <- table08 %>% 
  # filter(!is.na(Name) & Name != "") %>%  # Remove rows with NA or empty names
  filter(Name %in% c(paste0("B15001_00", seq(01,09),"E"),paste0("B15001_0", seq(10,83),"E")))
# %>% 
#    mutate(Label = str_replace_all(Label,", GED, or alternative", ' (includes equivalency)'))

#get column names  2012
url12 <- "https://api.census.gov/data/2012/acs/acs5/groups/B15001.html"

webpage12 <- read_html(url12)

table12 <- webpage12 %>%
  html_node("table") %>%  # Adjust the selector if necessary
  html_table() %>% 
  select(c("Name","Label","Concept","Required","Attributes","Limit","Predicate Type","Group"))

filteredtable12 <- table12 %>% 
  # filter(!is.na(Name) & Name != "") %>%  # Remove rows with NA or empty names
  filter(Name %in% c(paste0("B15001_00", seq(01,09),"E"),paste0("B15001_0", seq(10,83),"E")))
# %>% 
#    mutate(Label = str_replace_all(Label,", GED, or alternative", ' (includes equivalency)'))

#get column names 2016
url16 <- "https://api.census.gov/data/2016/acs/acs5/groups/B15001.html"

webpage16 <- read_html(url16)

table16 <- webpage16 %>%
  html_node("table") %>%  # Adjust the selector if necessary
  html_table() %>% 
  select(c("Name","Label","Concept","Required","Attributes","Limit","Predicate Type","Group"))

filteredtable16 <- table16 %>% 
  # filter(!is.na(Name) & Name != "") %>%  # Remove rows with NA or empty names
  filter(Name %in% c(paste0("B15001_00", seq(01,09),"E"),paste0("B15001_0", seq(10,83),"E"))) 

#get columnn names 2020
url20 <- "https://api.census.gov/data/2020/acs/acs5/groups/B15001.html"

webpage20 <- read_html(url20)

table20 <- webpage20 %>%
  html_node("table") %>%  # Adjust the selector if necessary
  html_table() %>% 
  select(c("Name","Label","Concept","Required","Attributes","Limit","Predicate Type","Group"))

filteredtable20 <- table20 %>% 
  # filter(!is.na(Name) & Name != "") %>%  # Remove rows with NA or empty names
  filter(Name %in% c(paste0("B15001_00", seq(01,09),"E"),paste0("B15001_0", seq(10,83),"E"))) %>% 
  mutate(Label = str_replace_all(Label,":",""))

```

```{r include=FALSE, results='hide'}
#Are column names the same across all ACS data?

filteredtable08==filteredtable12

filteredtable08==filteredtable16

filteredtable08==filteredtable20

filteredtable12==filteredtable16

filteredtable12==filteredtable20

filteredtable16==filteredtable20
```

```{r}
#update the mismatches
filteredtable08 <- filteredtable08 %>% 
   mutate(Label = str_replace_all(Label,", GED, or alternative", ' (includes equivalency)'))

filteredtable12 <- filteredtable12 %>% 
  mutate(Label = str_replace_all(Label,", GED, or alternative", ' (includes equivalency)'))
```

```{r include=FALSE, results='hide'}
#recheck 
#Are column names the same across all ACS data?

filteredtable08==filteredtable12

filteredtable08==filteredtable16

filteredtable08==filteredtable20

filteredtable12==filteredtable16

filteredtable12==filteredtable20

filteredtable16==filteredtable20

#yes, they are now so we can just use the column names from the latest ACS
```

All column names are the same across all 4 election year Educational Attainment data.

#### Combine and merge education data
```{r}
ed_attain <- rbind(ed_attain2008, ed_attain2012, ed_attain2016, ed_attain2020)

```

```{r}
ed_colnames <- filteredtable20 %>% 
  mutate(Name = str_replace_all(Name,"E","")) %>% 
  select(c(Name, Label))

sort(unique(ed_colnames$Name))==sort(unique(ed_attain$variable))

ed_attain2 <- left_join(ed_attain, ed_colnames, by = c("variable"="Name"))

glimpse(ed_attain2)
```

```{r}
#identify empty and NA values
colSums(ed_attain2 == "" | is.na(ed_attain2))
```
#### Clean and reshape data

```{r}
ed_attain3 <- ed_attain2 %>% 
  mutate(Label2 = Label,
         #Label2 = str_replace(Label2, "Estimate!!Total!!", ""),
         #Label2 = str_replace_all(Label2, " ","_")
         ) %>%
  separate(Label2, into = c("type","value","gender", "age_group", "education"), sep = "!!")
```

```{r}
head(ed_attain3, 10)
```

```{r}
#identify empty and NA values
colSums(ed_attain3 == "" | is.na(ed_attain3))
```


```{r}
ed_attain3_na <- ed_attain3 %>% 
  filter(is.na(gender) | is.na(age_group) | is.na(education))

ed_attain3_na %>% 
  count(variable, Label)

unique(ed_attain3_na$variable) 

# need to remove these from the whole data set and save as another
# reshape both data sets
# merge back together
```

```{r}

```
  
  
Other references and notes  
  
   
Esri prompt  
Because voting is voluntary in the United States, the level of voter participation
(referred to as "voter turnout") has a significant impact on the election results
and resulting public policy.  
Modeling voter turnout, and understanding where low turnout is prevalent, can
inform outreach efforts to increase voter participation. With the ultimate goal
of predicting voter turnout, in this exercise, you will focus on performing various
data engineering tasks to prepare election result data for predictive analysis.


Voter turnout  
https://electionlab.mit.edu/research/voter-turnout  
https://election.lab.ufl.edu/ (VEP- Voting Eligible Population) data by state and election, by county not available  
  
Redistricting   
https://www.propublica.org/article/redistricting-a-devils-dictionary (definitions)  
  
Class bias  
https://web.archive.org/web/20141107053600/http://academic.udayton.edu/grantneeley/pol%20303/avery%20and%20peffley%20-%20SPPQ%202005.pdf  
https://web.archive.org/web/20160412151014/https://www.nyu.edu/gsas/dept/politics/faculty/nagler/apsa2006_rv7.pdf  
  
Felon disenfranchisement  
https://web.archive.org/web/20190516191656/https://felonvoting.procon.org/sourcefiles/uggen_manza_asr_02.pdf  


<!------- Below is for removing excessive space in Rmarkdown | HTML formatting -------->

<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>